---
title: "Simple Linear Regression Example"
author: "Dr. Luis Malpica, modified from a stats in R course by Dr. Brett Favaro"
date: "April, 2020"
output:
  pdf_document:
    toc: yes
  html_document:
    theme: united
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Introduction
We will run a simple linear regression model on a dataset on tern wreckages due to hurricanes off SE USA, raw data available from: https://peerj.com/articles/3287/

Main variables are:
- wrecks = # of dead birds in a given two-week period, this will be our response (Y) variable
- hurricanes = # of hurricanes in a two week period, and this our predictive (X) variable

More details on model output interpretation is on the powerpoint presentation provided in a pdf file for the 11th week of the course.

# Install required packages & R custom made functions
```{r}
# If you don't yet have the following packages run:
# install.packages("tidyverse")
library(tidyverse)

# Load custom R functions
source("D:/UABC/II_Semestre/Datos_R/week11/functions/6003Functions.R")

```

# Load & clean up data steps
```{r}
# Read in data
terns <- read.csv("./data/6003_terns.csv")

# Remember the clean up data steps
# STEP 1: Did it load correctly?
# STEP 2: Check data types
# STEP 3: Check for missing or impossible values
# STEP 4: Check for typos and broken factors

# Skip data verification because dataset is very simple
head(terns)

# y =  wrecks = dead birds in a given two-week period
# x = hurricanes = # of hurricanes in a two week period

# Also, remember the "8 steps of Data Exploration", check for:
# 1. Outliers Y and X
# 2. Homogeneity Y
# 3. Normality Y
# 4. Zero trouble Y
# 5. Collinearity X('s)
# 6. Relationships Y and X
# 7. Interactions X('s)
# 8. Independence Y

# We will skip this part also...but do this for every dataset you work with as well

```

# Fit linear model to data (X as continuous variable)
```{r}
fit <- lm(wrecks ~ hurricanes, data=terns)

# Extract predicted and residuals values
terns$predicted <- predict(fit)
terns$residuals <- residuals(fit)

# Basic scatterplot
a <- ggplot(terns, aes(x = hurricanes, y = wrecks)) +
  geom_point() + theme_bw() 
a

# Add a lm

a + geom_smooth(method="lm", se=FALSE, color="darkgrey")

# Without lm, but with predicted values 

a + geom_point(aes(y=predicted), shape=1, color="blue")

# With predicted values + residuals

a + geom_point(aes(y=predicted), shape=1, color="blue") +
  geom_segment(aes(xend=hurricanes, yend=predicted), alpha=0.5, color="red")

# With line and residuals

a + geom_segment(aes(xend=hurricanes, yend=predicted), alpha=0.5, color="red") +
  geom_smooth(method="lm", se=FALSE, color="blue")

# With residuals visualized as actual squares

x1 <- terns$hurricanes
x2 <- terns$hurricanes+terns$residuals
y1 <- terns$predicted
y2 <- terns$predicted+terns$residuals

a + geom_smooth(method="lm", se=FALSE, color="blue") +
  geom_rect(aes(xmin=x1, ymin=y1, xmax=x2, ymax=y2), alpha=0.5)

```

# Model diagnostics & output exploration
```{r}
# Explore model diagnostics
# We look for homocedasticity and normality in residuals 
par(mfrow = c(2,2)) #this tells R to set up a 2x2 grid of 4 plots
plot(fit) #create the diagnostic plots
par(mfrow = c(1,1)) #turn off the 4x4 grid setup or all your future plots will go onto a grid

# Explore residuals distribution with a histogram
hist(residuals(fit), breaks=10)
# Residuals are normal-ish distributed but homocedasticity is not very well fulfilled

# Let's ignore these big concern and look at model output
summary(fit)
# Look at model output interpretation in powerpoint presentation

```

# Model visualization
```{r}
# Visualize model & data points
a <- ggplot(data=terns, aes(x=hurricanes, y=wrecks)) +
  theme_bw() +
  geom_point() +
  geom_smooth(method="lm", se=FALSE, color="darkgrey") 
a

# Add CIs
# First estimate them based on our fitted model
CIs <- predict(fit, interval="confidence", level = 0.95)

lwr <- CIs[,2] # Lower CI
upr <- CIs[,3] # Upper CI

# Option 1: ggplot2 calculates CIs
a + geom_smooth(method=lm, se=TRUE) #se = confidence interval

# Option 2: Use our CI's estimated above
a + geom_line(aes(y=upr, x=terns$hurricanes), col="red") + 
  geom_line(aes(y=lwr, x=terns$hurricanes), col="red") 

# You could also add the regression equation on the plot
# Plot model, points, and an equation for the line
a <- ggplot(terns, aes(x = hurricanes, y = wrecks)) + 
  geom_point() +
  theme_bw() +
  geom_smooth(method="lm", se=FALSE, color="darkgrey") +
  geom_line(aes(y=upr, x=terns$hurricanes), col="red") + 
  geom_line(aes(y=lwr, x=terns$hurricanes), col="red") +
  geom_text(x = 8, y = 23, label=lm_eqn(fit), parse=TRUE, size=6) 
a
```

# Model simulation
Here we will only estimate or predict datapoints based on our fitted model, this will help to illustrate the "inner workings of linear models". This next code estimates a normal distribution around 12 datapoints, each of these distributions is estimated based on the parameters of our fitted model. You do not need to run this for every model you make, it is just for visualization purposes.
```{r}

# Simple visualization of the model
plot(wrecks ~ hurricanes, 
      data=terns, pch=16,
     ylim=c(-10, 25))
abline(fit, col="darkgrey", lwd=2)


# Simulate predicted values and add to plot
md <- seq(0, 25, length = 12)
Beta <- coef(fit)
for (i in 1:12){
  mu <- Beta[1] + Beta[2] * md[i]
  yi <- rnorm(50, mean = mu, sd = summary(fit)$sigma)
  points(x = jitter(rep(md[i], 50)), 
         y = jitter(yi), 
         col = grey(0.5), 
         pch = 16, 
         cex = 1)
}
abline(h = 0, lty=2)

```

# Fit Linear Model to data (X as categorical variable)
Now let's turn hurricanes into a categorical variable, we could for example create group levels based on the number of hurricanes
```{r}
# We can use the median to divide our categorical variable into two groups
median(terns$hurricanes) 

terns2 <- terns %>%
  mutate(numstorms = ifelse(hurricanes < 7, "Level 1", "Level 2"))

terns2$numstorms <- as.factor(terns2$numstorms) # We need these variable as a factor

plot(wrecks ~ numstorms, data=terns2)#, xlab="Number of storms (categorical)")

categorical_lm1 <- lm(wrecks ~ numstorms, data=terns2)
summary(categorical_lm1)

# This is just like running a t-test
t.test(terns2$wrecks ~ terns2$numstorms)

#-------------------------------------------#
# We can further divide our categorical variable into three or more groups

terns3 <- terns %>%
  mutate(numstorms = ifelse(hurricanes < 5, "Level 1 (<5)", 
                            ifelse(hurricanes < 13, "Level 2 (6-12)", "Level 3 (14-23)")))

terns3$numstorms <- as.factor(terns3$numstorms)

plot(wrecks~numstorms, 
     data=terns3, xlab="Number of storms (categorical)")

categorical_lm <- lm(wrecks ~ numstorms, data=terns3)

summary(categorical_lm)

# This is just like running an ANOVA
anova_version <- aov(wrecks ~ numstorms, data=terns3)
summary(anova_version)

```

